Sample-efficient Reinforcement Learning with Latent Representations

### 2024-01-29 paper/main.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CUSTOM (our stuff)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Latin
\usepackage{xspace}
\newcommand{\eg}{\textit{e.g.\@}\xspace}
\newcommand{\ie}{\textit{i.e.\@}\xspace}
\newcommand{\cf}{\textit{cf.\@}\xspace}
\newcommand{\etc}{\textit{etc.\@}\xspace}
\newcommand{\etal}{\textit{et~al.\@}\xspace}

@@

\todo{Add algorithm}

\begin{align}
z_{t} &= e_{\theta} (s_{t}) \\
\hat{z}_{t+1} &= d_{\phi} (z_{t}, a_{t}) \\
q_{t} &= Q_{\psi} (z_{t}, a_{t}) \\
a_{t} &\sim \pi_{\eta} (z_{t}),
\end{align}

\textbf{Representation learning}
\begin{align}
  \mathcal{L}(\theta, \phi; \mathcal{D})
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
\end{align}

@@

@@

@@

\section{Theoretical Analysis}
% \label{}

Effect of domain shift on policy due to encoder being learned online

\citet{zhangLearningInvariantRepresentations2020} have proofs for bounds on value functions with encoder.
We could probably start with some of these.
They build off the idea of bisimulation \citep{larsenBisimulationProbabilisticTesting1989}.
Where encoders group states into clusters whilst preserving properties
(\eg optimal value, all values, all action values from each state) \citep{liUnifiedTheoryState2006}.

\textbf{Representation learning improves sample efficiency in RL}
% \begin{theorem}
% \label{thm:bigtheorem}
% Given two MDPs $\left( \mathcal{S}_{1}, \mathcal{A}, P, r, \rho_{0}, \gamma \right)$ and $\left( \mathcal{S}_{2}, \mathcal{A}, P, r, \rho_{0}, \gamma \right)$
% with $|\mathcal{S}_{1}| < |\mathcal{S}_{2}|$,
% the optimal value function $V : \mathcal{S} \rightarrow \R$ in $\mathcal{S}_{1}$ can be learned with less environment samples than in $\mathcal{S}_{2}$.
% \end{theorem}

% \begin{assumption}
% The state space $\mathcal{S}$ is discrete
% \label{ass:discrete-state-space}
% \end{assumption}

\begin{itemize}
    \item Prove latent space improves sample efficiency
    \begin{itemize}
        \item Assume discrete state space
        \item Assume latent space contains all information from state space needed to predict values
        \item Show smaller state space can be learned with less samples
    \end{itemize}
    \item Prove retraining from scratch is better than continuing training
    \begin{itemize}
        \item Training encoder episode to episode is like non-IID training
    \end{itemize}
\end{itemize}

@@

You can have as much text here as you want. The main body must be at most $8$ pages long.
For the final version, one more page can be added.
If you want, you can use an appendix like this one.

The $\mathtt{\backslash onecolumn}$ command above can be kept in place if you prefer a one-column appendix, or can be removed if you prefer a two-column appendix.  Apart from this possible change, the style (font size, spacing, margins, page numbering, etc.) should be kept the same as the main body.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CUSTOM (our stuff)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\our}{\textsc{vq-td3}\xspace}

% Latin
\usepackage{xspace}
\newcommand{\eg}{\textit{e.g.\@}\xspace}
\newcommand{\ie}{\textit{i.e.\@}\xspace}
\newcommand{\cf}{\textit{cf.\@}\xspace}
\newcommand{\etc}{\textit{etc.\@}\xspace}
\newcommand{\etal}{\textit{et~al.\@}\xspace}

@@

In this section, we detail our method, titled Vector Quantized TD3 (\our).
\our is conceptually simple, it first learns a representation of the observation space and then,
given this representation, it performs model-free RL (e.g. TD3) on this representation.
% It the first representation learning phase, it learns an encoder which maps observations $o \in \mathcal{O}$
% to a latent space $z \in \mathcal{Z}$.
% It first learns a representation of the observation space and then uses
See \cref{alg:main_alg}.

We consider Markov Decision Processes (MDPs, \citet{bellmanMarkovianDecisionProcess1957a}) $\mathcal{M} = (\mathcal{O}, \mathcal{A}, \mathcal{P}, \mathcal{R}, \gamma,\mathcal{T})$, where an agent receives
observation $o_{t} \in \mathcal{O}$ at time step $t$ and performs an action $a_{t} \in \mathcal{A}$.
The agent then obtains a reward $r_{t} = \mathcal{R} (o_{t}, a_{t})$ and obtains the next observation
$o_{t+1} = P(\cdot \mid o_{t}, a_{t})$.
discount factor $\gamma \in [0, 1]$

\our has four parameterized components.
\begin{algorithm}[tb]
   \caption{\our}
   \label{alg:main_alg}
\begin{algorithmic}
   \STATE {\bfseries Input:} Encoder $e_{\theta}$, transition $d_{\phi}$, critics $Q_{\psi}$, policy $\pi_{\eta}$
   \FOR{$i$ {\bfseries to} $N_{\text{episodes}}$}
    \STATE $\mathcal{D} \leftarrow \mathcal{D} \cup \{o_{t}, a_{t}, o_{t+1}, r_{t+1}\}^{T}_{t=0}$
    \FOR{$i=1$ {\bfseries to} $T \times UTD$}
        \STATE Representation update
        \STATE TD3 update
    \ENDFOR
   \ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{align}
\text{Encoder: } &z_{t} = e_{\theta} (s_{t}) \label{eq:encoder} \\
\text{Dynamics: } &\hat{z}_{t+1} = d_{\phi} (z_{t}, a_{t}) \label{eq:transition} \\
\text{Value: } &q_{t} = Q_{\psi} (z_{t}, a_{t}) \label{eq:value} \\
\text{Policy: } &a_{t} \sim \pi_{\eta} (z_{t}) , \label{eq:policy}
\end{align}
Note that unlike TCRL \citep{zhaoSimplifiedTemporalConsistency2023}, our transition model does
not predict the reward.

\textbf{Representation learning}
Our representation learning is based soley on the temporal consistency loss,
However, unlike previous works, we normalize the latent vector
Our representation learning consists of an encoder $e_{\theta}$ (which maps observations to
latent states) and a transition model $d_{\phi}$ which maps
Our representation learning adopts a self-supervised loss.

\begin{align} \label{eq:rep-loss}
  \mathcal{L}_{\text{rep}}(\theta, \phi; \mathcal{D})
 % &= \E_{s_{t} \sim \mathcal{D}} \left[ \sum_{h=0}^{H-1} \| \hat{z}_{t+1} - z_{t+1} \|_{2}^{2} \right] \\
%&= \E_{(s_{t}, a_{t}, s_{t+1}) \sim \mathcal{D}}
&= \E_{\tau \sim \mathcal{D}}
\left[ \sum_{h=0}^{H-1} \gamma^{h} \| d_{\phi}(e_{\theta}(s_{t}), a_{t}) - e_{\bar{\theta}}(s_{t+1}) \|_{2}^{2} \right]
\end{align}

\textbf{Finite Scalar Quantization}
We follow Finite Scalar Quantization (FSQ, \cite{mentzerFiniteScalarQuantization2023})
and bound each dimension of our latent space.
Given the output of our MLP encoder $z \in \R^{d}$,

where each entry in  $f : z \rightarrow \left[ L/2 \right] \mathrm{tanh}(z))$

% Thereby, we have zˆ ∈ C, where C is the implied codebook, given by the product of these per-channel
% codebook sets, with |C| = L
% d
% . The vectors in C can simply be enumerated leading to a bijection from
% any zˆ to an integer in {1, . . . , Ld}. Therefore, VQ can be replaced with FSQ in any neural networkrelated setup where VQ is commonly used, e.g., to train transformers, after appropriately adapting
% the output and input dimension of the layers before and after VQ, respectively. We generalize the
% above exposition to the case where the i-th channel is mapped to Li values and get |C| =
% Qd
% i=1 Li

@@

\begin{align} \label{eq:value-loss}
  \mathcal{L}_{Q}(\psi ; \mathcal{D}) &= \E_{\tau \sim \mathcal{D}} \left[ (Q_{\psi_{k}}(e_{\theta}(s_{t}), a_{t}) - y)^{2}  \right], \quad \forall k \in 1, 2 \\
  y &= \sum_{n=0}^{N-1} \gamma^{n} r_{t+n} + \gamma^{n} \min_{k \in \{1,2\}} Q_{\bar{\psi}_{k}}(e_{\bar{\theta}}(s_{t+n}), a_{t+n})
\end{align}

\begin{align} \label{eq:policy-loss}
 \mathcal{L}_{\pi}(\eta; \mathcal{D}) = \E_{(o_{t}, a_{t}) \sim \mathcal{D}} \left[ \min_{k\in\{1,2\}} Q_{\psi_{k}}(e_{\theta}(o_{t}), a_{t}) \right]
\end{align}

@@

\begin{figure*}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=1.0\textwidth]{./figs/main_plot.pdf}}
\caption{Comparison to baselines with UTD=1.}
\label{fig:main_plot}
\end{center}
\vskip -0.2in
\end{figure*}

@@

@@

\begin{figure*}[ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=0.97\textwidth]{./figs/all_mujoco_envs.pdf}}
\caption{Comparison to SAC/TCRL with UTD=1.}
\label{fig:main_plot}
\end{center}
\vskip -0.2in
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
